Sparse spectral and $p$-finite element methods on disk-slices and trapeziums I: zero boundary conditions

### 2019-04-03 papers/halfdisk/halfdisk.tex
Added explanation for general case of constructing the OPs to get 4 param family of 2D OPs

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, let  \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:

\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}

We consider the solution of partial differential equations on the domain
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}.
\end{align}
For simplicity of discussion we focus on the half-disk, where $\rho(x) = \sqrt{1-x^2}$,  $(\alpha,\beta) = (0, 1)$, and  $(\gamma, \delta)  = (-1,1)$, with discussion of extension to other geometries in the appendix.

@@

By using a similar process to \cite[p55--56]{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\).

@@

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= w_H^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}

@@

The work in this paper on the half disk can be easily transferred to the domain of a disk slice by which we mean
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
with 
\begin{align}
\begin{cases}
-1 \le \alpha < &\beta \le 1 \\
(\gamma, \delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w^{(a,b,c)}_1(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c} = (\beta - x)^a \: (x - \alpha)^{b} \: (1-x^2)^{c} \\
w^{(b)}_2(x) &:= (1-x^2)^b = (1-x)^b \: (1+x)^b.
\end{cases}
\end{align}
The weight $w^{(b)}_2(x)$ is a still an ultraspherical weight, and the corresponding OPs are the Jacobi polynomials $\{P_n^{(b, b)}\}$. $w^{(a,b,c)}_1(x)$ is the (non-classical) weight for the OPs denoted $\{H_n^{(a, b,c)}\}\). Thus we arrive at the two parameter family of 2D orthogonal polynomials $\{\Pnk\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
	\Pnk^{(a,b,c)}(x,y) := H_{n-k}^{(c, a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal with respect to the weight
\begin{align}
	W^{(a,b,c)}(x,y) &:= w^{(c,a,b)}_1(x) w^{(b)}_2(\frac{y}{\rho(x)}) \nonumber \\
	&= (\beta - x)^c \: (x - \alpha)^{a} \: (\rho(x)^2 -y^2)^b \nonumber \\
	&= (\beta - x)^c \: (x - \alpha)^{a} \: (1 - x^2 -y^2)^b , \quad (x,y) \in \Omega.
\end{align}

Note that by setting $c = 0$, $\alpha = 0$ and $\beta = 1$ we obtain the 2D orthogonal polynomials on the half-disk.

@@

We can further extend this work to trapezium shaped domains. Note that for any trapezium there exists an affine map to the canonical trapezium domain that we consider here, given by
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
with 
\begin{align}
\begin{cases}
(\alpha, \beta) &:= (0,1) \\
(\gamma, \delta) &:= (0,1) \\
\rho(x) &:= 1-\half x \\
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} = (1-x)^{a} \: x^b \: (1-\half x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b = (1-x)^a \: x^b.
\end{cases}
\end{align}
The weight $w^{(a,b)}_P(x)$ is the weight for the shifted Jacobi polynomials on the interval $[0,1]$, and hence the corresponding OPs are the shifted Jacobi polynomials $\{\tilde{P}_n^{(a, b)}\}$. We note that the shifted Jacobi polynomials relate the the normal Jacobi polynomials by the relationship $\tilde{P}_n^{(a,b)}(x) = P_n^{(a,b)}(2x-1)$ for any degree $n = 0,1,2,\dots$ and $x \in [0,1]$. $w^{(a,b,c)}_H(x)$ is the (non-classical) weight for the OPs we dentote $\{H_n^{(a,b,c)}\}\). Thus we arrive at the two parameter family of 2D orthogonal polynomials $\{\Pnk\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
	\Pnk^{(a,b,c,d)}(x,y) := H_{n-k}^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \tilde{P}_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal with respect to the weight
\begin{align}
	W^{(a,b,c,d)}(x,y) &:= w^{(a,b,c+d)}_H(x) \: w^{(d,c)}_P(\frac{y}{\rho(x)}) \nonumber \\
	&= (1-x)^a \: x^b \: y^c \: (1-\half x - y)^d, \quad (x,y) \in \Omega,
\end{align}
i.e. that
\begin{align}
	\int \int_\Omega \Pnk^{(a,b,c,d)} \: P_{m,j}^{(a,b,c,d)} \: W^{(a,b,c,d)} \: \dx \dy &:= C_{n,k}^{(a,b,c,d)} \: \delta_{n,m}
 \: \delta_{j,k},
\end{align}
for some constant $C_{n,k}^{(a,b,c,d)}$.

>>>>

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
where we require that either of the following scenarios hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and the weight function \(w_2\) on the interval $(\gamma, \delta)$ is an even function.
\end{enumerate}
We will explain where $w_2$ comes into play in Section \ref{Section:OPs}.

For simplicity of discussion we focus in this paper on the half-disk, where $\rho(x) = \sqrt{1-x^2}$,  $(\alpha,\beta) = (0, 1)$, and  $(\gamma, \delta)  = (-1,1)$, with discussion of extension to other geometries in the appendix.

@@

By using a similar process to \cite[p55--56]{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\). We generalise this process to demonstrate a method of constructing multi-parameter families of 2D orthogonal polynomials on disk-slice and trapezium shaped domains $\Omega$ from multi-parameter families of 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\).

@@

We can generalise this for our purposes in the following proposition.
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $w^{(a,b,c)}_H(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{H_n^{(a,b,c)}\}$ are orthogonal with respect to $w^{(a,b,c)}_H(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
Then the four-parameter 2D polynomials defined on $\Omega$ given by
\begin{align}
	\Pnk^{(a,b,c,d)}(x,y) := H_{n-k}^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
are orthogonal polynomials with respect to the weight 
\begin{align}
	W^{(a,b,c,d)}(x,y) := w^{(a,b,c+d)}_H(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align}
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= w_H^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
Note here the we have made the adjustment that $w^{(a,b,c)}_H(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

@@

The work in this paper on the half disk can be easily transferred to the domain of a disk slice by which we mean
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
with 
\begin{align}
\begin{cases}
(\alpha, \beta) &\subset (-1,1) \\
(\gamma, \delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half}.
\end{cases}
\end{align}
Our 1D weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively are then given by:
\begin{align}
\begin{cases}
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a)}_P(x) &:= (1-x^2)^a.
\end{cases}
\end{align}

The weight $w^{(a)}_P(x)$ is a still an ultraspherical weight, and the corresponding OPs are the Jacobi polynomials $\{P_n^{(a, a)}\}$. $w^{(a,b,c)}_H(x)$ is the (non-classical) weight for the OPs denoted $\{H_n^{(a,b,c)}\}\). Thus we arrive at the three-parameter family of 2D orthogonal polynomials $\{\Pnk^{(a,b,c)}\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
	\Pnk^{(a,b,c)}(x,y) := H_{n-k}^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: P_k^{(c,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal with respect to the weight
\begin{align}
	W^{(a,b,c)}(x,y) &:= w^{(a,b,2c)}_1(x) w^{(c)}_2(\frac{y}{\rho(x)}) \nonumber \\
	&= (\beta - x)^a \: (x - \alpha)^{b} \: (\rho(x)^2 -y^2)^c \nonumber \\
	&= (\beta - x)^a \: (x - \alpha)^{b} \: (1 - x^2 -y^2)^c , \quad (x,y) \in \Omega.
\end{align}

We can note that by setting the parameter $c = 0$, and $\alpha = 0$, $\beta = 1$ we obtain the 2D orthogonal polynomials on the half-disk (note again also that we made the adjustment that $w^{(a,b,c)}_H(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ for the half-disk case, for ease of writing and computation).

@@

We can further extend this work to trapezium shaped domains. Note that for any trapezium there exists an affine map to the canonical trapezium domain that we consider here, given by
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
with 
\begin{align}
\begin{cases}
(\alpha, \beta) &:= (0,1) \\
(\gamma, \delta) &:= (0,1) \\
\rho(x) &:= 1-\half x \\
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} = (1-x)^{a} \: x^b \: (1-\half x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b = (1-x)^a \: x^b.
\end{cases}
\end{align}
The weight $w^{(a,b)}_P(x)$ is the weight for the shifted Jacobi polynomials on the interval $[0,1]$, and hence the corresponding OPs are the shifted Jacobi polynomials $\{\tilde{P}_n^{(a, b)}\}$. We note that the shifted Jacobi polynomials relate the the normal Jacobi polynomials by the relationship $\tilde{P}_n^{(a,b)}(x) = P_n^{(a,b)}(2x-1)$ for any degree $n = 0,1,2,\dots$ and $x \in [0,1]$. $w^{(a,b,c)}_H(x)$ is the (non-classical) weight for the OPs we dentote $\{H_n^{(a,b,c)}\}\). Thus we arrive at the four-parameter family of 2D orthogonal polynomials $\{\Pnk^{(a,b,c,d)}\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
	\Pnk^{(a,b,c,d)}(x,y) := H_{n-k}^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \tilde{P}_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal with respect to the weight
\begin{align}
	W^{(a,b,c,d)}(x,y) &:= w^{(a,b,c+d)}_H(x) \: w^{(d,c)}_P(\frac{y}{\rho(x)}) \nonumber \\
	&= (1-x)^a \: x^b \: y^c \: (1-\half x - y)^d, \quad (x,y) \in \Omega.
\end{align}
